---
title: "Index"
author: "Jian Che"
date: "8/5/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### *Problem of Domain & Description of variables*

For this dataset I want to look at the trend of prison population in each state
and their proportion of the country. I also want to explore the racial disparities
in prison, which race is being admitted into prison more, and whether this 
disparity lowered or increased in recent years. Where the imprisonment occured
also relate to racial disparities since many suburban areas are largely gentrified.
I want to look at which area have a higher imprisonment number and if there
is a common similarity among all states.
```{r include=FALSE}
# load libraries
library(stringr)
library(dplyr)
library(ggplot2)
# library(tidyverse)
library(hrbrthemes)
library(viridis)
setwd('~/info201/a3---data-visualization-and-applications-staccjch/incarceration-trends')

# Load datasets
jailpop_year_df <- read.csv('year-end-prison-2021.csv')
incar_trend_df <- read.csv('incarceration_trends.csv')
```

### *Summary Values*

* State with Highest Prison Population 2019-2021: Texas
  + Texas 2019: 158820
  + Texas 2020: 136302
  + Texas 2021: 133424
* State with the MOST Population Change:
  + Alaska: 159
* State with the LEAST Population Change:
  + Texas: -25329
* Which state has the highest Black adult imprisonment in 2018?
  + Texas: 2439277
* Which state has the highest Latin adult imprisonment in 2018? 
  + California: 10477829
* hich state has the highest female adult imprisonment in 2018? 
  + California: 13092077


```{r include=FALSE}
# Which state has the highest jail population each year?
jail_pop_by_state <- jailpop_year_df[8:57, 2:5]
max_2019_state <- jail_pop_by_state[jail_pop_by_state$total_prison_pop_2019
                                   == max(jail_pop_by_state$total_prison_pop_2019), 
                                   'state_name', drop = FALSE]
max_2020_state <- jail_pop_by_state[jail_pop_by_state$total_prison_pop_2020
                                   == max(jail_pop_by_state$total_prison_pop_2020), 
                                   'state_name', drop = FALSE]
max_2021_state <- jail_pop_by_state[jail_pop_by_state$total_prison_pop_2021
                                    == max(jail_pop_by_state$total_prison_pop_2021), 
                                    'state_name', drop = FALSE]
# All Texas
# Which state has the max difference from 2019-2021
jail_pop_by_state <- jail_pop_by_state %>%
  mutate(pop_growth = total_prison_pop_2021 - total_prison_pop_2019)

max_growth <- jail_pop_by_state[jail_pop_by_state$pop_growth
                                    == max(jail_pop_by_state$pop_growth), 
                                    'state_name', drop = FALSE]
print(max_growth)  # Alaska = 159

min_growth <- jail_pop_by_state[jail_pop_by_state$pop_growth
                                == min(jail_pop_by_state$pop_growth), 
                                'state_name', drop = FALSE]
print(min_growth) # Texas = -25369

# Information in 2018
incar_trend_2018 <- incar_trend_df %>%
  filter(year == 2018) %>%
  select(state, total_pop:male_pop_15to64, aapi_pop_15to64:white_pop_15to64) %>%
  group_by(state) %>%
  summarise_each(funs(sum))
# Which state has the highest Black adult imprisonment in 2018? Texas = 2439277
max_black_2018 <- incar_trend_2018[incar_trend_2018$black_pop_15to64
                                   == max(incar_trend_2018$black_pop_15to64), 
                                   'state', drop = FALSE]

# Which state has the highest Latin adult imprisonment in 2018? CA = 10477829
max_Latin_2018 <- incar_trend_2018[incar_trend_2018$latinx_pop_15to64
                                   == max(incar_trend_2018$latinx_pop_15to64), 
                                   'state', drop = FALSE]
# Which state has the highest female adult imprisonment in 2018? CA = 13092077
max_female_2018 <- incar_trend_2018[incar_trend_2018$female_pop_15to64
                                   == max(incar_trend_2018$female_pop_15to64), 
                                   'state', drop = FALSE]
```

## Trends over time chart
I included the chart of top 10 Black Imprisonment States's incarceration trend from 1990 to 2018. This chart illustrated a high amount of incarceration trend on Black, showing trend of racism. While some States have lowered their Black incarceration rate over time, like New York and Illinois, other States like Texas and California, Florida rised over time. 

```{r echo=FALSE}
# library(hrbrthemes)
# library(viridis)
# Trend over time chart: top 10 states Black imprisonment population over 1990-2018
# First get top 10 states names
top_10_black_prison <- incar_trend_2018 %>%
  select(state, black_pop_15to64) %>%
  arrange(desc(black_pop_15to64)) %>%
  slice(1:10)
top_10_black_vector <- top_10_black_prison $state
print(top_10_black_vector)
# filter out top 10 Black imprisonment states' data
black_trend_df <- incar_trend_df %>%
  filter(year%in% 1990:2018) %>%
  select(state, year, county_name, black_pop_15to64)
black_trend_df <- black_trend_df[black_trend_df$state %in% top_10_black_vector, ]
# get the sum of all counties
Black <- aggregate(black_pop_15to64~state+year, data = black_trend_df, FUN = sum)

# Plot
Black %>%
  ggplot( aes(x=year, y=black_pop_15to64, group=state, color=state)) +
  geom_line() +
  scale_color_viridis(discrete = TRUE) +
  ggtitle("Top 10 Black Imprisonment States Incarceration Trend") +
  theme_ipsum() +
  ylab("Black imprisonment population")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Varaible Comparison chart
My variable comparison chart showed the relevance between Black imprisonment population and jail rate capacity. The graph showed a general positive linear correlation between jail rate capacity 
and Black prisoner population. However, there are also a clump of scatterpoints at the low end of 
jail rate capacity, indicating that even the smaller jails have a exceedingly high proportion of
black prisoners. This shows racial disparities because Black prisoners take up a lot of the jail's 
capacity regardless of jail sizes. 
```{r include=FALSE}
# Variable Comparison Chart: Relationship between jail rated capacity and black admissions rate

# # plotly Plot
library(reshape2)
library(plotly)
library(tidymodels)
# library(hrbrthemes)
# library(viridis)
```
```{r echo=FALSE}
jail_capacity_df <- incar_trend_df %>%
  select(year, state, black_prison_pop, jail_rated_capacity) %>%
  na.omit()
x <- jail_capacity_df$jail_rated_capacity
y <- jail_capacity_df$black_prison_pop

lm_model <- linear_reg() %>%
  set_engine('lm') %>%
  set_mode('regression') %>%
  fit(jail_rated_capacity ~ black_prison_pop, data = jail_capacity_df) 

fit <- lm(jail_rated_capacity ~ black_prison_pop, data = jail_capacity_df)

jail_capacity_df %>%
  plot_ly(x = ~jail_rated_capacity) %>%
  add_markers( y = ~black_prison_pop) 
  # %>% add_lines(x = ~jail_rated_capacity, y = fitted(fit))

admissions <- incar_trend_df %>%
  select(year, state, fips, county_name, total_prison_adm) %>%
  na.omit()
admissions_2016 <- admissions %>% filter(year == 2016)
#admissions <- aggregate(total_prison_adm~state+year, data = admissions, FUN = sum)
```


## Map
My map showed the total prison admissions population in 2016 in each US counties. Through this map, we can see which State and County have higher imprisonment rate, which might also implies the crime rate trend of the country in general.It shows racial disparities such that counties with high 
prison admissions rate are also less gentrified. 
```{r echo=FALSE}
library(usmap)
# library(hrbrthemes)
# library(viridis)
# library(ggplot2)
plot_usmap(data = admissions_2016, values = 'total_prison_adm', color = 'grey') +
  scale_fill_continuous(low = 'lightblue2', high = 'navy', name = 'prison admission(2016)', 
                        label = scales::comma) +
  labs(title = 'Counties Prison Admissions', subtitle = 'This map shows prison admission in 2016') +
  theme(legend.position = "right")
```